Inference for multivariate peaks over threshold models

### 2021-10-30 methodology_projgamma.tex
updates

\subsubsection{Inference for the projected Gamma mixture model}
The Dirichlet process mixture model groups observations into stochastically assigned \emph{clusters}.  Building out the methods of inference for Equation~\ref{eqn:dppg}, let $n_j^{-i}$ be the number of observations in cluster $j$ not including observation $i$.  Let $J^{-i}$ be the number of extant observations, not including any singleton that observation $i$ is in. Under this model, the probability of cluster membership is proportional to
\begin{equation*}
    \text{P}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
        n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\hspace{0.5cm} &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
        \eta\int\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)dG_0(\bm{\alpha}_j,\bm{\beta}_j)\hspace{0.5cm} &\text{~for~}j = J^{-i} + 1.
        \end{cases}
\end{equation*}
Of course, if $G_0$ is not a conjugate prior for the kernel distribution, then the integral is difficult.  Instead, using algorithm 8 from \cite{neal2000}, we draw $m$ candidate clusters, $\bm{\alpha}_j,\bm{\beta}_j$ for $j = J^{-i} + 1,\ldots, J^{-i} + m$ from $G_0$. Then, the probability of cluster membership is proportional to
\begin{equation}
    \text{P}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
        n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\hspace{0.5cm} &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
        \frac{\eta}{m}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\hspace{0.5cm} &\text{~for~}j \in \lbrace J^{-i} + 1,\ldots, J^{-i} + m\rbrace.
        \end{cases}
\end{equation}
Of course, if a candidate cluster is selected, then $\delta_j = J + 1$, and $J = J^{-i} + 1$.

A key feature of the the projected Gamma distribution is its computational properties.  With respect to the distributional parameters $\bm{\alpha}_i,\bm{\beta}_i$, the projected Gamma density is proportional to
\begin{equation*}
    \mathcal{PG}(\bm{y}\mid\bm{\alpha}_i,\bm{\beta}_i) \propto \int\prod_{\ell = 1}^d\mathcal{G}\left(ry_{i\ell}\mid\alpha_{i\ell},\beta_{i\ell}\right)r^{d-1}dr,
\end{equation*}
which implies that, by reintroducing the latent radial component $r_i$ via sampling from its full conditional,
\begin{equation}
    r_i\mid\bm{\alpha}_i,\bm{\beta}_i \sim \mathcal{G}\left(\sum_{\ell = 1}^d\alpha_{i\ell}, \sum_{\ell = 1}^d\beta_{\ell} y_{i\ell}\right),
\end{equation}
the likelihood for $\bm{\alpha},\bm{\beta}$ becomes separable by dimension.  We stress this point: \emph{given} $r$, the distributional parameters $(\alpha_j,\beta_j)_{\ell}$ are independent in the likelihood.

The full conditional for $\bm{\alpha}_j,\bm{\beta}_j$ is then proportional to
\begin{equation}
    f(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{Y},\bm{r},\bm{\delta},\ldots) \propto \prod_{i:\gamma_i = j}\prod_{\ell = 1}^d\mathcal{G}\left(r_iy_{i\ell}\mid\alpha_{j\ell},\beta_{j\ell}\right) \times dG_0(\alpha_{\ell},\beta_{\ell})
\end{equation}
The first form of the centering distribution we explore is a product of independent Gammas.
\begin{equation}
    G_0(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{\xi},\bm{\tau},\bm{\zeta},\bm{\sigma}) = \prod_{\ell = 1}^d\mathcal{G}(\alpha_{j\ell}\mid \xi_{\ell},\tau_{\ell})\times\prod_{\ell = 2}^d\mathcal{G}(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell})
\end{equation}
This model is completed with Gamma priors on $\xi_{\ell}$, $\tau_{\ell}$, $\zeta_{\ell}$, $\sigma_{\ell}$.  Optionally, placing a Gamma prior on $\eta$ allows updating via the procedure outlined in \cite{escobar1995}.  In further analysis, we refer to this model as the \emph{projected Gamma--Gamma} (PG--G) model.  An advantage to using a Gamma centering distribution for the rate parameter is, as a conjugate prior, the rate parameters can easily be integrated out for inference on $\bm{\alpha}_j$.  After doing so, the full conditional for $\alpha_{j\ell}$ takes the form
\begin{equation}
    \pi(\alpha_{j\ell}\mid \bm{r},\bm{Y},\bm{\gamma},\xi_\ell,\tau_\ell,\zeta_\ell,\sigma_\ell) \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i\ell}\right)^{\alpha_{j\ell} - 1}\alpha_{j\ell}^{\xi_\ell - 1}e^{-\tau_\ell \alpha_{j\ell}}\times\Gamma\left(n_j\alpha_{j\ell} + \zeta_{\ell}\right)}{\Gamma^{n_j}(\alpha_{j\ell})\left(\sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right)^{n_j\alpha_{j\ell} + \zeta_{\ell}}}
\end{equation}
for $\ell = 2,\ldots,d$.  For $\ell = 1$, as $\beta_{1} := 1$, the full conditional takes the form
\begin{equation}
    \label{eqn:alpha1update}
    \pi(\alpha_{j1}\mid\bm{r},\bm{Y},\bm{\gamma},\xi_1,\tau_1) \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i1}\right)^{\alpha_{j1} - 1}\alpha_{j1}^{\zeta_1 - 1}e^{-\tau_1\alpha_{j1}}}{\Gamma^{n_j}(\alpha_{j1})}.
\end{equation}
Updating these parameters can be done via a simple Metropolis-within-Gibbs step.  In our analysis, we first transform $\alpha_{j\ell}$ to log scale, and use a normal proposal density.
The full conditional for $\beta$ is then
\begin{equation}
    \beta_{j\ell}\mid\bm{r},\bm{Y},\alpha,\zeta_{\ell},\sigma_{\ell} \sim \mathcal{G}\left(\beta_{j\ell}\mid n_j\alpha_{j\ell} + \zeta_\ell, \sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right),
\end{equation}
for $\ell = 2,\ldots, d$.  Updating $\beta_{j\ell}$ is done via a Gibbs step.  The hyper-parameters $\xi_{\ell},\tau_{\ell},\zeta_{\ell},\sigma_{\ell}$ follow similar Gamma-Gamma update relationships.  We also explore a restricted form of this model, where $\beta_{\ell} = 1$ for all $\ell$.  Under this model, we use the full conditional in Equation~\ref{eqn:alpha1update} for all $\ell$, and omit inference on $\bm{\zeta},\bm{\sigma}$.  We refer to this model as the \emph{projected restricted Gamma--Gamma} (PRG--G) model.

>>>>

\subsubsection{Inference for the projected Gamma mixture model}
The Dirichlet process mixture model groups observations into stochastically assigned \emph{clusters}.  Building out the methods of inference for Equation~\ref{eqn:dppg}, let $n_j^{-i}$ be the number of observations in cluster $j$ not including observation $i$.  Let $J^{-i}$ be the number of extant observations, not including any singleton that observation $i$ is in. Under this model, the probability of cluster membership is proportional to
\begin{equation*}
    \text{P}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
        n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\hspace{0.5cm} &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
        \eta\int\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)dG_0(\bm{\alpha}_j,\bm{\beta}_j)\hspace{0.5cm} &\text{~for~}j = J^{-i} + 1.
        \end{cases}
\end{equation*}
Of course, if $G_0$ is not a conjugate prior for the kernel distribution, then the integral is difficult.  Instead, using algorithm 8 from \cite{neal2000}, we draw $m$ candidate clusters, $\bm{\alpha}_j,\bm{\beta}_j$ for $j = J^{-i} + 1,\ldots, J^{-i} + m$ from $G_0$. Then, the probability of cluster membership is proportional to
\begin{equation}
    \text{P}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
        n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\hspace{0.5cm} &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
        \frac{\eta}{m}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\hspace{0.5cm} &\text{~for~}j \in \lbrace J^{-i} + 1,\ldots, J^{-i} + m\rbrace.
        \end{cases}
\end{equation}
Of course, if a candidate cluster is selected, then $\delta_j = J = J^{- i} + 1$, and the associated cluster parameters are saved.

A key feature of the the projected Gamma distribution is its computational properties.  With respect to the distributional parameters $\bm{\alpha}_i,\bm{\beta}_i$, the projected Gamma density is proportional to
\begin{equation*}
    \mathcal{PG}(\bm{y}\mid\bm{\alpha}_i,\bm{\beta}_i) = \int\prod_{\ell = 1}^d\mathcal{G}\left(ry_{i\ell}\mid\alpha_{i\ell},\beta_{i\ell}\right)r^{d-1}\left[y_d +
            {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(y_d^p\right)^{\frac{1}{p} - 1}\right]dr,
\end{equation*}
which implies that, by reintroducing the latent radial component $r_i$ via sampling from its full conditional,
\begin{equation}
    r_i\mid\bm{\alpha}_i,\bm{\beta}_i \sim \mathcal{G}\left(\sum_{\ell = 1}^d\alpha_{i\ell}, \sum_{\ell = 1}^d\beta_{\ell} y_{i\ell}\right),
\end{equation}
the likelihood for $\bm{\alpha},\bm{\beta}$ becomes separable by dimension.  We stress this point: \emph{given} $r$, the dimensions of the distributional parameters $\bm{\alpha}_j,\bm{\beta}_j$ are independent in the likelihood.

The full conditional for $\bm{\alpha}_j,\bm{\beta}_j$ is then proportional to
\begin{equation}
    f(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{Y},\bm{r},\bm{\delta},\ldots) \propto \prod_{i:\gamma_i = j}\prod_{\ell = 1}^d\mathcal{G}\left(r_iy_{i\ell}\mid\alpha_{j\ell},\beta_{j\ell}\right) \times dG_0(\alpha_{\ell},\beta_{\ell})
\end{equation}
The first form of the centering distribution we explore is a product of independent Gammas:
\begin{equation}
    G_0(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{\xi},\bm{\tau},\bm{\zeta},\bm{\sigma}) = \prod_{\ell = 1}^d\mathcal{G}(\alpha_{j\ell}\mid \xi_{\ell},\tau_{\ell})\times\prod_{\ell = 2}^d\mathcal{G}(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}).
\end{equation}
This model is completed with Gamma priors on $\xi_{\ell}$, $\tau_{\ell}$, $\zeta_{\ell}$, $\sigma_{\ell}$.  We also placed a Gamma prior on $\eta$, updating via the procedure outlined in \cite{escobar1995}.  In further analysis, we refer to this model as the \emph{projected Gamma--Gamma} (PG--G) model.  An advantage to using a Gamma centering distribution for the rate parameter is, as a conjugate prior, the rate parameters can easily be integrated out for inference on $\bm{\alpha}_j$.  After doing so, the full conditional for $\alpha_{j\ell}$ takes the form
\begin{equation}
    \pi(\alpha_{j\ell}\mid \bm{r},\bm{Y},\bm{\gamma},\xi_\ell,\tau_\ell,\zeta_\ell,\sigma_\ell) \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i\ell}\right)^{\alpha_{j\ell} - 1}\alpha_{j\ell}^{\xi_\ell - 1}e^{-\tau_\ell \alpha_{j\ell}}\times\Gamma\left(n_j\alpha_{j\ell} + \zeta_{\ell}\right)}{\Gamma^{n_j}(\alpha_{j\ell})\left(\sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right)^{n_j\alpha_{j\ell} + \zeta_{\ell}}}
\end{equation}
for $\ell = 2,\ldots,d$.  For $\ell = 1$, as $\beta_{1} := 1$, the full conditional takes the form
\begin{equation}
    \label{eqn:alpha1update}
    \pi(\alpha_{j1}\mid\bm{r},\bm{Y},\bm{\gamma},\xi_1,\tau_1) \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i1}\right)^{\alpha_{j1} - 1}\alpha_{j1}^{\zeta_1 - 1}e^{-\tau_1\alpha_{j1}}}{\Gamma^{n_j}(\alpha_{j1})}.
\end{equation}
Updating these parameters can be done via a simple Metropolis-within-Gibbs step.  In our analysis, we first transform $\alpha_{j\ell}$ to log scale, and use a normal proposal density.
The full conditional for $\beta$ is then
\begin{equation}
    \beta_{j\ell}\mid\bm{r},\bm{Y},\alpha,\zeta_{\ell},\sigma_{\ell} \sim \mathcal{G}\left(\beta_{j\ell}\mid n_j\alpha_{j\ell} + \zeta_\ell, \sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right),
\end{equation}
for $\ell = 2,\ldots, d$.  Updating $\beta_{j\ell}$ is done via a Gibbs step.  The hyper-parameters $\xi_{\ell},\tau_{\ell},\zeta_{\ell},\sigma_{\ell}$ follow similar Gamma-Gamma update relationships.  We also explore a restricted form of this model, where $\beta_{\ell} = 1$ for all $\ell$.  Under this model, we use the full conditional in Equation~\ref{eqn:alpha1update} for all $\ell$, and omit inference on $\bm{\zeta},\bm{\sigma}$.  We refer to this model as the \emph{projected restricted Gamma--Gamma} (PRG--G) model.

### 2021-10-30 results.tex
updates

\begin{figure}[ht]
  \caption{Simulation Energy Score versus Column Count, by Mixture Component Count. Lower is better.\label{fig:simes}}
  \centering
  \includegraphics[width=0.8\textwidth]{./images/simulation_es}
\end{figure}

@@

\begin{figure}[ht]
  \caption{Simulation posterior predictive loss versus column count, by mixture component count.\label{fig:simppl}}
  \centering
  \includegraphics[width=0.8\textwidth]{./images/simulation_ppl}
\end{figure}

@@

\subsection{Integrated Vapor Transport\label{subsec:ivt}}
The \emph{integrated vapor transport} (IVT) tracks the total water volume in a column of air over a given
  area \citep{ralph2017}.  IVT is the means by which we track the formation and path of atmospheric 
  rivers---elongated areas of high local concentration of water vapor in the atmosphere that moves with
  wind patterns.  Extreme behavior in IVT can lead to extreme precipitation events, causing flooding; 
  but in California, those same events are responsible for the bulk of precipitation that inland 
  California recieves---water necessary for agriculture and household consumption.  Thus it becomes 
  necessary to manage precipitation to prevent inundation and support agriculture.  To this end, 
  understanding the dependence structure of extreme events in the IVT is useful---for instance, for 
  hydrological engineers in planning the capacity of groundwater systems to capture precipitation before 
  wasteful and destructive flooding occurs.  Our data tracks daily average values for the IVT, dividing 
  the coast of California into grid cells, with two spatial resolutions of 8 and 47 cells, respectively 
  \citep{guan2015}.

Fitting our models to this data requires some pre-processing. First, we subset the data to the rainy
  season, which appears from November to March.  The marginal distributions of the
  IVT data appear naturally log-normal, which falls into the domain of attraction of a Gumbel
  distribution.  Given that, we can apply a high threshold, and exceedances over that threshold can
  be modelled as Generalized Pareto.  Our modelling begins with that assessment.  As estimating
  the Pareto parameters is not yet our focus in this analysis, we choose to apply the threshold
  using the empirical CDF.  That is, for a given $t$, let $b_{t,\ell} = \hat{F}_{\ell}^{-1}(1 - t^{-1})$.  For
  this analysis, we set $t = 20$, indicating the marginal $95$ percentile.  The other parameters of the
  generalized Pareto---the scale parameter $\alpha_{\ell}$ and the extremal index $\chi_{\ell}$---are set via
  maximum likelihood.  A fully Bayesian model formulation will allow their varying within a
  distribution, but such a model formulation will not be conducive to fitting by Markov-chain Monte
  Carlo methods.

After the thresholding and maximum likelihood estimation of the parameters of the Pareto, we scale
  the data to the standard multivariate Pareto.  Dividing each standardized observation by its
  $\mathcal{L}_{\infty}$ norm, we project the standardized data onto $\mathcal{S}_{\infty}^{d-1}$.
  Data in sequence represents observations in time, and these are heavily correlated.  As such, we
  choose to \emph{decluster} the observations, such that, by observing a sequence of observations
  $\bm{ z}$ for which $\inorm{z_i} > 1$ for each observation in sequence, we keep only the observation
  with the greatest observed $\mathcal{L}_{\infty}$ norm.  The complete procedure is outlined in
  Algorithm~\ref{algo:processing}.

\begin{algorithm}
  \KwResult{$\bm{ r},\bm{v} : r_i \sim \text{Pareto}(1)$, $\bm{ v}_i \in \mathcal{S}_{\infty}^{d-1}$}
  \For{$\ell = 1,\ldots,d$}{
    Set $b_{t,\ell} = \hat{F}_{\ell}^{-1}\left(1 - \frac{1}{t}\right)$.\\
    With $\bm{ x}_{\ell} > b_{t,\ell}$, fit $a_{\ell}$, $\chi_{\ell}$ via MLE according to generalized Pareto likelihood.\\
    }
  \For{$i = 1,\ldots,n$}{
    Define $z_{i,\ell} = \left(1 + \xi_{\ell}\frac{x_{i,\ell} - b_{t,\ell}}{a_{\ell}}\right)_{+}^{1/\xi_{\ell}}$\\
    Define $r_i = \pnorm{\bm{ z}_i}{\infty}$, $\bm{ v}_i = \frac{\bm{ z}_i}{\pnorm{\bm{ z}_i}{\infty}}$\\
    }
  Subset $\bm{ r},\bm{ v}$ such that $r_i \geq 1$ for all $r_i\in \bm{r}$.\\
  \If{declustering}{
    \For{$i = 1,\ldots,n$}{
      If $r_i \geq 1$ and $r_{i-1} \geq 1$, drop the lesser (and associated $v_i$) from data set.\\
    }
  }
 \caption{Data preprocessing to isolate and transform data exhibiting extreme behavior.  $r_i$
   represents the radial component, and $\bm{v}_i$ the angular component.  The declustering
   portion is relevant if data is correlated in time.\label{algo:processing}}
\end{algorithm}

We have, at our disposal, two data sets from the IVT model, totalling approximately 30 years of daily
  values omitting leap days.  One records data from 8 grid cells, covering
  generally the coast of California.  The other, with a higher resolution, records data from 47 grid
  cells covering the same area.  For the 8-cell data, after subsetting to the rainy season, we have
  $5587$ observations.  After the data processing steps in Algorithm~\ref{algo:processing}, where we
  omit observations not extreme in any dimension, and decluster temporally correlated data, we are left
  with $511$ observations.  For the 47-cell data, after subsetting to the rainy season, we have $6342$
  observations.  After the same data processing, we are left with
  $532$ observations.  We fit our models to both data sets.

@@

In Table~\ref{tab:dev} we see the same model preference towards the projected restricted gamma models
  that we saw in Figures~\ref{fig:simes}.  The unrestricted gamma model is penalized much
  more strongly on real data than we saw with our simulation, performing comparably to the probit-normal
  model.  We also see the restricted gamma model with the log-normal prior performing significantly
  better than the restricted gamma model with the gamma prior on the higher dimensional data, a reversal
  of what we saw on the low dimensional data and indeed from what we saw in the simulation study.

>>>>

% \begin{figure}[ht]
%   \caption{Simulation Energy Score versus Column Count, by Mixture Component Count. Lower is better.\label{fig:simes}}
%   \centering
%   \includegraphics[width=0.8\textwidth]{./images/simulation_es}
% \end{figure}

\begin{figure}[ht]
  \begin{subfigure}[b]{0.45\textwidth}
    \includegraphics[width=\textwidth]{./images/simulation_ppl}
    \caption{Posterior Predictive Loss\label{fig:simppl}}
  \end{subfigure}
  %
  \begin{subfigure}[b]{0.45\textwidth}
    \includegraphics[width=\textwidth]{./images/simulation_es}
    \caption{Energy Score\label{fig:simes}}
  \end{subfigure}
  \caption{Simulation PPL (left) and ES (left) versus column count, by mixture component count.}
\end{figure}

@@

% \begin{figure}[ht]
%   \caption{Simulation posterior predictive loss versus column count, by mixture component count.\label{fig:simppl}}
%   \centering
%   \includegraphics[width=0.8\textwidth]{./images/simulation_ppl}
% \end{figure}

@@

\subsection{Integrated Vapor Transport\label{subsec:ivt}}
The \emph{integrated vapor transport} (IVT) tracks the total water volume in a column of air over a given
  area \citep{ralph2017}.  IVT is the means by which we track the formation and path of atmospheric 
  rivers---elongated areas of high local concentration of water vapor in the atmosphere that moves with
  wind patterns.  Extreme behavior in IVT can lead to extreme precipitation events, causing flooding; 
  but in California, those same events are responsible for the bulk of precipitation that inland 
  California receives---water necessary for agriculture and household consumption.  Thus it becomes 
  necessary to manage precipitation to prevent inundation and support agriculture.  To this end, 
  understanding the dependence structure of extreme events in the IVT is useful---for instance, for 
  hydrological engineers in planning the capacity of groundwater systems to capture precipitation before 
  wasteful and destructive flooding occurs.  Our data tracks daily average values for the IVT, dividing 
  the coast of California into grid cells, with two spatial resolutions of 8 and 47 cells, respectively 
  \citep{guan2015}.

\begin{algorithm}[h]
  \KwResult{$\bm{ r},\bm{v} : r_i \sim \text{Pareto}(1)$, $\bm{ v}_i \in {\mathbb S}_{\infty}^{d-1}$}
  \For{$\ell = 1,\ldots,d$}{
    Set $b_{t,\ell} = \hat{F}_{\ell}^{-1}\left(1 - \frac{1}{t}\right)$.\\
    With $\bm{ x}_{\ell} > b_{t,\ell}$, fit $a_{\ell}$, $\chi_{\ell}$ via MLE according to generalized Pareto likelihood.\\
    }
  \For{$i = 1,\ldots,n$}{
    Define $z_{i,\ell} = \left(1 + \xi_{\ell}\frac{x_{i,\ell} - b_{t,\ell}}{a_{\ell}}\right)_{+}^{1/\xi_{\ell}}$\\
    Define $r_i = \pnorm{\bm{ z}_i}{\infty}$, $\bm{ v}_i = \frac{\bm{ z}_i}{\pnorm{\bm{ z}_i}{\infty}}$\\
    }
  Subset $\bm{ r},\bm{ v}$ such that $r_i \geq 1$\\
  \If{declustering}{
    \For{$i = 1,\ldots,n$}{
      If $r_i \geq 1$ and $r_{i-1} \geq 1$, drop the lesser (and associated $v_i$) from data set.\\
    }
  }
 \caption{Data preprocessing to isolate and transform data exhibiting extreme behavior.  $r_i$
   represents the radial component, and $\bm{v}_i$ the angular component.  The declustering
   portion is relevant if data is correlated in time.\label{algo:processing}}
\end{algorithm}

Fitting our models to this data requires some pre-processing. First, we subset the data to the rainy
  season, which appears from November to March.  The marginal distributions of the
  IVT data appear naturally log-normal, which falls into the domain of attraction of a Gumbel
  distribution.  Given that, we can apply a high threshold, and exceedances over that threshold can
  be modelled as Generalized Pareto.  Our modelling begins with that assessment.  As estimating
  the Pareto parameters is not yet our focus in this analysis, we choose to apply the threshold
  using the empirical CDF.  That is, for a given $t$, let $b_{t,\ell} = \hat{F}_{\ell}^{-1}(1 - t^{-1})$.  For
  this analysis, we set $t = 20$, indicating the marginal $95$ percentile.  The other parameters of the
  generalized Pareto---the scale parameter $\alpha_{\ell}$ and the extremal index $\chi_{\ell}$---are set via
  maximum likelihood.

After the thresholding and maximum likelihood estimation of the parameters of the Pareto, we scale
  the data to the standard multivariate Pareto following Equation~\ref{eqn:standardization}.  
  Dividing each standardized observation by its $\mathcal{L}_{\infty}$ norm, we project the 
  standardized data onto $\mathcal{S}_{\infty}^{d-1}$. Data in sequence represents observations
  in time, which for this application are temporally correlated.  As such, we
  choose to \emph{decluster} the observations, such that, by observing a sequence of observations
  $\bm{ z}$ for which $\inorm{z_i} > 1$ for each observation in sequence, we keep only that observation
  with the greatest observed $\mathcal{L}_{\infty}$ norm.  The complete procedure is outlined in
  Algorithm~\ref{algo:processing}.  In the 8-cell data, after subsetting to the rainy season we 
  have $5587$ observations, which reduces to $511$ observations after processing and declustering.  
  In the 47 cell data, after subsetting we have $6342$ observations, which reduces to $532$ after 
  processing and declustering.  We fit PG--G, PRG--G, PG--LN, and PRG--LN models to both datasets.

@@

In Table~\ref{tab:dev} we see the same model preference towards the projected restricted 
  gamma models that we saw in Figures~\ref{fig:simes}.  The unrestricted gamma model is 
  penalized much more strongly on real data than we saw with our simulation.  We also see
  the restricted gamma model with the log-normal prior performing significantly better 
  than the restricted gamma model with the gamma prior on the higher dimensional data, a reversal
  of what we saw on the low dimensional data and indeed from what we saw in the simulation study.

\begin{figure}[ht!]
    \centering
    \caption{Pairwise extremal dependence coefficients for IVT data, from fitted PRG--G.  
    The left corresponds to the 8-cell data, the right the 47-cell data.\label{fig:chi_ij}}
    \begin{minipage}{.49\textwidth}
      \centering
      \includegraphics[width=0.99\linewidth]{./images/chi_ij_8}
    \end{minipage}
    \begin{minipage}{.49\textwidth}
      \centering
      \includegraphics[width=0.99\linewidth]{./images/chi_ij_46}
    \end{minipage}
\end{figure}

Figure~\ref{fig:chi_ij} shows our implementation of Equation~\ref{eqn:chi_ij}, plotting the 
  pairwise extremal dependence coefficients for the IVT data computed from PRG--G models 
  fitted to the 8 (left) and 47 (right) cell data.  
  That we see a higher extremal dependence between neighboring columns is not unexpected, as
  neighboring columns correspond to neighboring cells.  We recognize that pairwise asymptotic
  dependence coefficients tell a limited story, as a particular dependence may structure may include
  more than two columns.   We can, however, glean some information from the patterns that emerge in
  two dimensions.  On the 8-cell data, we see a stronger association between cells 5-8, indicating a
  greater dependence among these cells.  On the 47 cell data, we see at least 3 groupings of columns,
  indicating greater asymptotic dependence among these groups.
